A Meta-analysis of Longevity Estimates of Mosquito Vectors of Disease: Additional Methods

### 2018-11-29 MS.tex
Changes to remove major vector references

In Fig. \ref{fig:mrr_elpd}, we compare the performance of the six models for describing
lifespan in MRR studies of 33 species using K-fold cross-validation. We
categorised the evidence for age-dependent mortality in each species
according to the performance of the five age-dependent models versus the
exponential: `+' indicated that all age-dependent models outperformed
the exponential; `?' indicated that the exponential outperformed one or
more age-dependent models; and `-' indicated that the exponential
performed at least as well as all other models. Overall, we estimated
that there were 8 `+' species, where age-dependent mortality fit the
data better; 11 `?' species where the evidence was mixed; and 14 species
where constant mortality models performed at least as well. The species
where age-dependent mortality best fit the data included the major
vector of dengue fever, Zika and chikungunya, \emph{Ae. Aegypti}. These
studies also tended to include multiple release MRR studies which, on
average, were conducted over a longer period of time than the others,
which may be why we failed to detect age-dependence in the latter (Fig
\ref{S-fig:mrr_mcPowerAnalysis_senescence}).

@@

For malaria, estimates of the minimum fraction of the population that
can transmit the disease vary from \textless{}0.1\% for \emph{A.
subpictus} (posterior median; from the MRR analysis, as noted above likely to be due to
the LBL substantially underestimating lifespan) to 52\% (posterior median) for the
drought-adapted and long-lived \emph{A sergentii.} The proportions
surviving long enough to become infectious for the two major African
malaria vectors were: for \emph{A. gambiae s.l.}: 10\% (MRR) and 27\%
(dissection); and for \emph{A. funestus s.l.}: 9\% (MRR). Using the
individual time series estimates, there
evidence for a difference in EIP between the species (Kruskal-Wallis used due to non-normality of data; MRR:
\(\chi_{14}^{2} = 30.2\), \emph{p} \textless{}0.01; dissection: \(\chi_{11}^{2} = 38.9\), p\textless{}0.01).

@@

To compare estimates of lifespan derived from MRR with those from
dissection-based methods, we display the estimates of lifespan from
those ten species occurring in both databases in a single plot (Fig. \ref{fig:comparison}). In is reassuring that there is correlation between estimates from both approaches, although the small sample size likely hindered our ability
to determine statistical significance. In both cases, we estimate that
\emph{A. sergentii} was amongst the longest lived of the anopheline
species with an LBL of 12.4 days (mean estimate;
25\%-75\% CI: 5.9-13.8 days) from the MRR analysis and 11.9 days (mean
estimate; 25\%-50\% CI: 7.6-14.0 days) from the analysis of dissection
studies. This species is a major vector of malaria in the Sahara \citep{sinka2010dominant}, where to act as a disease vector it must persevere
through these hard conditions. It is reasonable to hypothesise that this
species should live longer than those in environments where the
potential for blood-feeding and oviposition is greater. The species with
the greatest discrepancy in the estimates was the major African malaria
vector \emph{A. gambiae s.l.}, where we estimated LBLs of 4.5 days
(mean estimate; 25\%-75\% CI: 3.8-5.1 days for unfed female) from the
MRR analysis and 9.5 days (mean estimate; 25\%-75\% CI: 5.2-11.0) from
the dissection analysis. Across genera, the greatest discrepancy in estimates was for
\emph{Aedes}, where the estimates from the MRR studies (8.1 days) are considerably longer than those of dissection-based studies (3.5 days).
This was followed by \emph{Culex} (a posterior mean of 2.9 days from the
MRR versus 4.9 days from the dissection analysis) with the smallest
discrepancy for \emph{Anopheles} (6.8 versus 6.4 days). Across all studies we estimate
from the MRR analysis that mean mosquito lifespan is 6.0 days versus 5.5
days from the dissection-based studies. Some of the differences in these
group-level estimates between the two approaches is likely due to
environmental and genetic differences between mosquitoes in the
experiments that were analysed in each meta-analysis. However, we
believe that part of the discrepancy can be explained by the
methodological differences in approaches. We speculate that differences
in dispersal rate can explain some of the discrepancy. Both
\emph{Anopheles} and \emph{Culex} mosquitoes are generally thought to
fly farther during their lifetimes than \emph{Aedes} {[}Charles, do you
have a reference here?{]}, meaning that the estimates from MRR-based
approaches will be most downwardly-biased for these genera. This is
supported by our results since the dissection-based estimates
(themselves not reliant on assumptions about dispersal) are similar or exceed the MRR
estimates for \emph{Anopheles} and \emph{Culex} mosquitoes, but not for
\emph{Aedes}.

>>>>

In Fig. \ref{fig:mrr_elpd}, we compare the performance of the six models for describing
lifespan in MRR studies of 33 species using K-fold cross-validation. We
categorised the evidence for age-dependent mortality in each species
according to the performance of the five age-dependent models versus the
exponential: `+' indicated that all age-dependent models outperformed
the exponential; `?' indicated that the exponential outperformed one or
more age-dependent models; and `-' indicated that the exponential
performed at least as well as all other models. Overall, we estimated
that there were 8 `+' species, where age-dependent mortality fit the
data better; 11 `?' species where the evidence was mixed; and 14 species
where constant mortality models performed at least as well. The species
where age-dependent mortality best fit the data included the
vector of dengue fever, Zika and chikungunya, \emph{Ae. Aegypti}. These
studies also tended to include multiple release MRR studies which, on
average, were conducted over a longer period of time than the others,
which may be why we failed to detect age-dependence in the latter (Fig
\ref{S-fig:mrr_mcPowerAnalysis_senescence}).

@@

For malaria, estimates of the minimum fraction of the population that
can transmit the disease vary from \textless{}0.1\% for \emph{A.
subpictus} (posterior median; from the MRR analysis, as noted above likely to be due to
the LBL substantially underestimating lifespan) to 52\% (posterior median) for the
drought-adapted and long-lived \emph{A sergentii.} The proportions
surviving long enough to become infectious for \emph{A. gambiae s.l.}: 10\% (MRR) and 27\%
(dissection); and for \emph{A. funestus s.l.}: 9\% (MRR). Using the
individual time series estimates, there
evidence for a difference in EIP between the species (Kruskal-Wallis used due to non-normality of data; MRR:
\(\chi_{14}^{2} = 30.2\), \emph{p} \textless{}0.01; dissection: \(\chi_{11}^{2} = 38.9\), p\textless{}0.01).

@@

To compare estimates of lifespan derived from MRR with those from
dissection-based methods, we display the estimates of lifespan from
those ten species occurring in both databases in a single plot (Fig. \ref{fig:comparison}). In is reassuring that there is correlation between estimates from both approaches, although the small sample size likely hindered our ability
to determine statistical significance. In both cases, we estimate that
\emph{A. sergentii} was amongst the longest lived of the anopheline
species with an LBL of 12.4 days (mean estimate;
25\%-75\% CI: 5.9-13.8 days) from the MRR analysis and 11.9 days (mean
estimate; 25\%-50\% CI: 7.6-14.0 days) from the analysis of dissection
studies. This species is a vector of malaria in the Sahara \citep{sinka2010dominant}, where to act as a disease vector it must persevere
through these hard conditions. It is reasonable to hypothesise that this
species should live longer than those in environments where the
potential for blood-feeding and oviposition is greater. The species with
the greatest discrepancy in the estimates was \emph{A. gambiae s.l.}, where we estimated LBLs of 4.5 days
(mean estimate; 25\%-75\% CI: 3.8-5.1 days for unfed female) from the
MRR analysis and 9.5 days (mean estimate; 25\%-75\% CI: 5.2-11.0) from
the dissection analysis. Across genera, the greatest discrepancy in estimates was for
\emph{Aedes}, where the estimates from the MRR studies (8.1 days) are considerably longer than those of dissection-based studies (3.5 days).
This was followed by \emph{Culex} (a posterior mean of 2.9 days from the
MRR versus 4.9 days from the dissection analysis) with the smallest
discrepancy for \emph{Anopheles} (6.8 versus 6.4 days). Across all studies we estimate
from the MRR analysis that mean mosquito lifespan is 6.0 days versus 5.5
days from the dissection-based studies. Some of the differences in these
group-level estimates between the two approaches is likely due to
environmental and genetic differences between mosquitoes in the
experiments that were analysed in each meta-analysis. However, we
believe that part of the discrepancy can be explained by the
methodological differences in approaches. We speculate that differences
in dispersal rate can explain some of the discrepancy. Both
\emph{Anopheles} and \emph{Culex} mosquitoes are generally thought to
fly farther during their lifetimes than \emph{Aedes} {[}Charles, do you
have a reference here?{]}, meaning that the estimates from MRR-based
approaches will be most downwardly-biased for these genera. This is
supported by our results since the dissection-based estimates
(themselves not reliant on assumptions about dispersal) are similar or exceed the MRR
estimates for \emph{Anopheles} and \emph{Culex} mosquitoes, but not for
\emph{Aedes}.
